---
title: "Efficient testing procedures for multiple time-to-event endpoints in group-sequential clinical trials"
subtitle: "A practical comparison based on simulated data"
author: "Moritz Fabian Danzer"
date: "last-modified"
bibliography: multsurvtrialdesign_bib.bib
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
    theme: cosmo
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
  error: false
  freeze: auto
knitr:
  opts_chunk:
    fig.path: "vignettes/figures/"
---

With this Markdown file, we provide a practical comparison of the multiple testing procedure introduced in @Danzer:2025 with already existing testing procedures. These procedures target settings with multiple primary time-to-event endpoints in a group-sequential clinical trial. While the method itself is generally applicable, we restrict ourselves to a setting with the two time-to-event endpoints progression-free survival (PFS) and overall survival (OS) and a trial design with one interim and one final analysis. The analysis cutoff dates are event-driven. As outlined in @Danzer:2025, both endpoints can be used to define these event-driven cutoffs. 

We require the following R packages:

* `simIDM` for simulating individual patient data (IPD) for the trial design considered here,
* `rpact` for calculating the critical values for the group-sequential testing procedures,
* `mvtnorm` for calculating multivariate normal probabilities,
* `tidyverse` for data manipulation.

The computation of test statistics, estimation of correlations between those test statistics and the calculation of inflation factors for the local significance levels is done by functions defined in the R script `stat_and_var_calc.R`.

```{r}
#| label: setup
# Load R packages
library(simIDM)
library(rpact)
library(mvtnorm)
library(tidyverse)

# Source R scripts with functions for the analyses
source("../R/stat_and_var_calc.R")
```

# Introduction

As in the manuscript, we consider eight (?) different testing procedure. These can be described as follows:

1. **BON**: Bonferroni-adjusted testing procedure with a test of the PFS hyopthesis at level $\rho_{\text{PFS}} \alpha$ at the interim analysis and a test of the OS hypothesis at level $\rho_{\text{OS}} \alpha$ at the final analysis.
2. **REC/LAST**: Bonferroni-adjusted testing procedure with a test of the PFS hyopthesis at level $\rho_{\text{PFS}} \alpha$ at the interim analysis and a test of the OS hypothesis at level $\rho_{\text{OS}} \alpha$ at the final analysis. However, in contrast to the BON procedure, the REC/LAST procedure allows for a reallocation of the alpha level from the PFS hypothesis to the OS hypothesis if the PFS hypothesis is rejected at the interim analysis. Specifically, if the PFS hypothesis is rejected at the interim analysis, then the alpha level for the OS hypothesis at the final analysis is increased to $\alpha$.
3. **REC/FIRST**: Bonferroni-adjusted testing procedure with a test of the PFS hyopthesis at level $\rho_{\text{PFS}} \alpha$ at the interim analysis and a test of the OS hypothesis at level $\rho_{\text{OS}} \alpha$ at the final analysis. However, in contrast to the BON procedure, the REC/FIRST procedure allows for a reallocation of the alpha level from the PFS hypothesis to the OS hypothesis if the PFS hypothesis is rejected at the interim analysis. Specifically, if the PFS hypothesis is rejected at the interim analysis, then the OS hypothesis is tested at level $\rho_{\text{PFS}} \alpha$ at the interim analysis. If unsuccessful, the remaining level $\rho_{\text{OS}} \alpha$ is spent at the final analysis as in a $\alpha$-spending approach.
4. **EX/LAST**: Improved closed testing procedure with exhaustion of the FWER. In addition to REC/LAST, we exploit the correlation of test statistics for PFS and OS to inflate the local level of the test of the OS hypothesis if the PFS hypothesis has not been rejected at the interim analysis.
5. **EX/FIRST**: Improved closed testing procedure with exhaustion of the FWER. In addition to REC/FIRST, we exploit the correlation of test statistics for PFS and OS to inflate the local level of the test of the OS hypothesis if the PFS hypothesis has not been rejected at the interim analysis.
6. **BON/GS**: Bonferroni-adjusted testing procedure with a test of the PFS hyopthesis at level $\rho_{\text{PFS}} \alpha$ at the interim analysis and a group-sequential procedure for the OS hypothesis at level $\rho_{\text{PFS}} \alpha$ according to an $\alpha$-spending approach based one the O'Brien-Fleming spending function of Lan and DeMets [see @Lan:83].  
7. **REC/GS/LAST**: Bonferroni-adjusted testing procedure with a test of the PFS hyopthesis at level $\rho_{\text{PFS}} \alpha$ at the interim analysis and a group-sequential procedure for the OS hypothesis at level $\rho_{\text{OS}} \alpha$ according to an $\alpha$-spending approach based one the O'Brien-Fleming spending function of Lan and DeMets [see @Lan:83]. As in the REC/LAST procedure, a rejection of the PFS hypothesis at the interim analysis allows to spend an additional alpha level of $\rho_{\text{PFS}} \alpha$ for the OS hypothesis at the final analysis.
8. **REC/GS/FIRST**: Bonferroni-adjusted testing procedure with a test of the PFS hyopthesis at level $\rho_{\text{PFS}} \alpha$ at the interim analysis and a group-sequential procedure for the OS hypothesis at level $\rho_{\text{OS}} \alpha$ according to an $\alpha$-spending approach based one the O'Brien-Fleming spending function of Lan and DeMets [see @Lan:83]. As in the REC/FIRST procedure, a rejection of the PFS hypothesis at the interim analysis allows to spend an additional alpha level of $\rho_{\text{PFS}} \alpha$ for the OS hypothesis at the interim analysis.
9. **EX/GS/LAST**: Improved closed testing procedure with exhaustion of the FWER. Based on REC/GS/LAST, we improve the procedure by inflating the local levels at the interim analysis by explotining the correlation of the test statistics for PFS and OS at the interim analysis. Additionally, if no hypothesis has been rejected at the interim analysis, we exploit the correlation of the OS test statistics at the final analysis with the two test statistics just mentioned.
10. **EX/GS/FIRST**: Improved closed testing procedure with exhaustion of the FWER. Based on REC/GS/FIRST, we improve the procedure by inflating the local levels at the interim analysis by explotining the correlation of the test statistics for PFS and OS at the interim analysis. Additionally, if no hypothesis has been rejected at the interim analysis, we exploit the correlation of the OS test statistics at the final analysis with the two test statistics just mentioned.
11. **OS**: Testing procedure with a test of the OS hypothesis at level $\alpha$ at the final analysis without any interim analysis.

Building on the very simple procedures BON and BON/GS, the procedures with the prefix "REC" allow for a reallocation of the significance level if the global null hypothesis has been rejected. The procedures with the prefix "EX" exploit the correlation of test statistics for PFS and OS. The resulting inflation of local significance level yields a uniform improvement of the testing procedures that takes effect as long as the global null hypothesis has not been rejected. 

# Trial design

The parameters of the Bonferroni correction on which all other procedures are based is given in the following code chunk.

```{r}
#| label: bonferroni_parameters
bretz_weight_pfs <- 1 / 5
bretz_weight_os <- 4 / 5
alpha <- 0.025
```

## Analysis dates

As mentioned above, we consider one interim and a final analysis. The cutoff date for the interim analysis is driven by the number of observed PFS events. The cutoff date for the final analysis is driven by the number of observed OS events. The number of events required for the interim and final analysis are given in the following code chunk.

```{r}
#| label: analysis_dates
nEventPFS <- 433
nEventOS <- 630
```

## Individual patient data for PFS and OS

Individual patient data for PFS and OS are simulated based on a time-homogeneous Markovian multi-state model as described e.g. in @Meller:2019. We simulate data with the `simIDM` package. We use the first scenario of @Erdmann:2025 here. The parameters for both treatment groups (i.e. the intensities for the transitions between the states of stable disease (0), progressive disease (1) and death (2)) are given in the follwoing code chunk.

```{r}
#| label: idm_settings
transitionTrt <- exponential_transition(h01 = 0.06, h02 = 0.3, h12 = 0.3)
transitionPl <- exponential_transition(h01 = 0.1, h02 = 0.4, h12 = 0.3)
transitionList <- list(transitionPl, transitionTrt)
```

## Recruitment and censoring mechanism

Patients are recruited over a time period of 32 months with a constant recruitment rate of 50 patients per month. As we assume a 1:1 randomisation ratio, this corresponds to a total sample size of 1600 patients with 800 patients in each treatment group. Besides administrative censoring at the end of the trial, we assume that there is additional loss to follow-up with a rate of 10% in 12 months.

```{r}
#| label: recruitment_censoring
# Recruitment
recr_rate_per_arm <- 25
recr_duration <- 32
max_n_per_arm <- 32 * 25

# Loss to follow-up
drprate <- 0.1
drptime <- 12
```

## Simulated data

All of the parameters described above are used to simulate individual patient data for PFS and OS for the trial design considered here. The simulated data are stored in the `Sim` object. The function `getDatasetWideFormat()` is used to transform the simulated data into a format that is suitable for the analyses of the testing procedures.

```{r}
#| label: simulate_pfs_os
set.seed(1262025)
Sim <- getDatasetWideFormat(getOneClinicalTrial(
  nPat = rep(max_n_per_arm, 2),
  transitionByArm = transitionList,
  dropout = list(rate = drprate, time = drptime),
  accrual = list(param = "intensity", value = recr_rate_per_arm)
))
```

# Interim analysis

## Application of administrative censoring 

To obtain the data to be analysed at the interim analysis, we apply administrative censoring to the simulated data. The first few rows of the resulting dataset are shown in the following code chunk.

```{r}
#| label: admin_censoring_interim
studyCensored1 <- censoringByNumberEvents(
  data = Sim,
  eventNum = nEventPFS,
  typeEvent = "PFS"
)
studyCensored1 <- studyCensored1 |>
  mutate(group = trt - 1) |>
  select(-c("CensoredOS", "trt")) |>
  relocate(group, .after = "id")
kable(head(studyCensored1))
```

## Computation of test statistics and estimated correlation matrix

We use our own function `test_stat_calc_single()` to compute the test statistics for the PFS and OS hypotheses at the interim analysis. The function also estimates the correlation between these test statistics.

```{r}
#| label: test_stats_interim
interim_stats <- test_stat_calc_single(studyCensored1)
interim_raw_stats <- interim_stats[[1]]
interim_var_est <- interim_stats[[2]]
# Compute standardised test statistics
interim_std_stats <- interim_raw_stats / sqrt(diag(interim_var_est))
```

The vector of interim test statistics $(U_{\text{PFS}}(A_{\text{PFS}}), U_{\text{OS}}(A_{\text{PFS}}))$, according to the notation of @Danzer:2025, is given by
$$
\begin{pmatrix}
  `{r} round(interim_raw_stats[1], 4)` \\
  `{r} round(interim_raw_stats[2], 4)`
\end{pmatrix}
$$

The estimated covariance matrix is given by
$$
\begin{pmatrix}
  `{r} round(interim_var_est[1, 1], 4)` & `{r} round(interim_var_est[1, 2], 4)` \\
  `{r} round(interim_var_est[2, 1], 4)` & `{r} round(interim_var_est[2, 2], 4)`
\end{pmatrix}
$$

The standardised test statistics thus amount to

$$
U_{\text{PFS}}(A_{\text{PFS}}) = `{r} round(interim_std_stats[1], 4)` \quad\text{ and }\quad U_{\text{OS}}(A_{\text{PFS}}) = `{r} round(interim_std_stats[2], 4)`.
$$

## Testing the global null hypothesis

For the 11 testing procedures mentioned above, we now want to outline how to determine the decisions at the interim analysis. We can reject $H_{0,\text{glob}}$ based on the test statistics $U_{\text{PFS}}(A_{\text{PFS}})$ and $U_{\text{OS}}(A_{\text{PFS}})$.

### Established procedures

We first consider the seven designs that do not explicitly exploit the correlation sctructure between the test statistics. For the group-sequential procedures, the local significance level for OS at the interim analysis is determined by the O'Brien-Fleming spending function of Lan and DeMets [see @Lan:83]. 
$$
f_{\text{OS, glob}}(t) = 2 - 2 \Phi\left(\frac{\Phi^{-1}(1 - \frac{\rho_{\text{OS}}\alpha}{2})}{\sqrt{t}}\right)
$$

Here, $t$ is the information fraction for OS at the interim analysis. It is implemented by the `getDesignGroupSequential()` function of the `rpact` package using the information fraction for OS at the interim analysis. Its computation and the resulting local significance level for OS at the interim analysis are given in the following code chunk. Using the notation of our manuscript, this information fraction is denoted by $\tau_{\text{OS}}(A_{\text{PFS}})$.

```{r}
#| label: local_os_levels_interim
info_fraction_os <- sum(studyCensored1$OSevent) / nEventOS
os_gs_design_bon <- getDesignGroupSequential(
  kMax = 2,
  alpha = alpha * bretz_weight_os,
  sided = 1,
  typeOfDesign = "asOF",
  informationRates = c(info_fraction_os, 1)
)
# Extract local levels of group-sequential design
os_stageLevels_bon <- os_gs_design_bon$stageLevels
# Extract cumulative alpha spent
os_alphaSpent_bon <- os_gs_design_bon$alphaSpent
# Extract alpha increments
os_alphaIncr_bon <- (c(
  os_alphaSpent_bon,
  alpha * bretz_weight_os
) -
  c(0, os_alphaSpent_bon))[1:2]
# Extract critical values
os_criticalValues_bon <- os_gs_design_bon$criticalValues
```

The following table contains the local significance levels for the test of the PFS and OS hypotheses at the interim analysis that would imply a rejection of the global null hypothesis.

| Procedure   | Local level for PFS | Critical value for PFS | Local level for OS | Critical value for OS|
|-----|-----:|-----:|-----:|-----:|
| BON         | `{r} alpha * bretz_weight_pfs` | `{r} round(qnorm(alpha * bretz_weight_pfs), 4)` | 0 | - |
| REC/LAST    | `{r} alpha * bretz_weight_pfs` | `{r} round(qnorm(alpha * bretz_weight_pfs), 4)` | 0 | - |
| REC/FIRST   | `{r} alpha * bretz_weight_pfs` | `{r} round(qnorm(alpha * bretz_weight_pfs), 4)` | 0 | - |
| BON/GS      | `{r} alpha * bretz_weight_pfs` | `{r} round(qnorm(alpha * bretz_weight_pfs), 4)` | `{r} round(os_stageLevels_bon[1], 4)` | `{r} round(qnorm(os_stageLevels_bon[1]), 4)` |
| REC/GS/LAST | `{r} alpha * bretz_weight_pfs` | `{r} round(qnorm(alpha * bretz_weight_pfs), 4)` | `{r} round(os_stageLevels_bon[1], 4)` | `{r} round(qnorm(os_stageLevels_bon[1]), 4)` |
| REC/GS/FIRST| `{r} alpha * bretz_weight_pfs` | `{r} round(qnorm(alpha * bretz_weight_pfs), 4)` | `{r} round(os_stageLevels_bon[1], 4)` | `{r} round(qnorm(os_stageLevels_bon[1]), 4)` |
| OS          | 0 | - | 0 | - |

The corresponding critical values have been computed by inserting these local levels $\alpha_{\text{loc}}$ into the standard normal quantile function $\Phi^{-1}$.  

### New procedures

For the newly proposed procedures EX/LAST, EX/FIRST, EX/GS/LAST and EX/GS/FIRST, we can exploit the correlation of the test statistics for PFS and OS at the interim analysis to inflate the local significance levels as long as the global null hypothesis has not been rejected. However, for EX/LAST and EX/FIRST, we cannot make an improvement on the established procedures because there is only one test statistic investigated at the interim analysis. Hence, we cannot exploit any correlation right now.

In comparison to REC/GS/LAST and REC/GS/FIRST, the new procedures EX/GS/LAST and EX/GS/FIRST use an inflation factor $\xi_1$ that is chosen in such a way that the total alpha spent for both test statistic is `{r} round(alpha * bretz_weight_pfs + os_alphaIncr_bon[1], 4)`. Of course, this computation requires the estimated covariance matrix displayed above. It can be carried out by the `inflation_factor()` function defined in the R script `stat_and_var_calc.R` as shown in the following code chunk.

```{r}
#| label: inflation_factor_interim
inflation_factor_interim <- inflation_factor(
  current_levels = c(alpha * bretz_weight_pfs, os_alphaIncr_bon[1]),
  available_level = alpha * bretz_weight_pfs + os_alphaIncr_bon[1],
  covariance_matrix = interim_var_est
)
```

With this inflation factor, the local levels for the new procedures amount to

| Procedure | Inflation factor| Local level for PFS | Critical value for PFS | Local level for OS | Critical value for OS|
|-----|-----:|-----:|-----:|-----:|-----:|
| EX/LAST | 1 |`{r} alpha * bretz_weight_pfs`| `{r} round(qnorm(alpha * bretz_weight_pfs), 4)` | 0 | - |
| EX/FIRST | 1 |`{r} alpha * bretz_weight_pfs`| `{r} round(qnorm(alpha * bretz_weight_pfs), 4)` | 0 | - |
| EX/GS/LAST | `{r} round(inflation_factor_interim, 4)`|`{r} round(inflation_factor_interim * alpha * bretz_weight_pfs, 4)` | `{r} round(qnorm(inflation_factor_interim * alpha * bretz_weight_pfs), 4)` | `{r} round(inflation_factor_interim * os_stageLevels_bon[1], 4)` | `{r} round(qnorm(inflation_factor_interim * os_stageLevels_bon[1]), 4)` |
| EX/GS/FIRST | `{r} round(inflation_factor_interim, 4)`|`{r} round(inflation_factor_interim * alpha * bretz_weight_pfs, 4)` | `{r} round(qnorm(inflation_factor_interim * alpha * bretz_weight_pfs), 4)` | `{r} round(inflation_factor_interim * os_stageLevels_bon[1], 4)` | `{r} round(qnorm(inflation_factor_interim * os_stageLevels_bon[1]), 4)` |

## Recycling the type I error level

As soon as the global null hypothesis has been rejected, we can reallocate the level from the endpoint that led to the rejection to the other endpoint. This can happen if $U_{\text{PFS}}(A_{\text{PFS}})$ or $U_{\text{OS}}(A_{\text{PFS}})$ exceed the critical values given above. If $U_{\text{PFS}}(A_{\text{PFS}})$ exceeds the critical value for PFS, then we can reallocate the level from PFS to OS. If $U_{\text{OS}}(A_{\text{PFS}})$ exceeds the critical value for OS, then we can reallocate the level from OS to PFS. The resulting local levels for the remaining test statistics.

### Reallocation from PFS to OS

If $U_{\text{PFS}}(A_{\text{PFS}})$ exceeds the critical bounds mentioned above, the local level is reallocated to the test of the OS hypothesis. The procedures we consider here deal differently with this reallocated level. The BON procedure does not use the reallocated level at all. The procedures with the suffix "LAST" use the reallocated level at the final analysis. Hence, the critocal values do not differ from those given above. The procedures with the suffix "FIRST" use the reallocated level directly at the interim analysis. This leads to different $\alpha$-spending functions. 

**REC/LAST and EX/LAST**:

Text
$$
f_{\text{OS, rec}}(t) =  \rho_{\text{PFS}}\alpha \cdot 1(t \geq 1) + \rho_{\text{OS}}\alpha \cdot 1(t \geq 1) = \alpha \cdot 1(t \geq 1)
$$

As we can seem no alpha is spent at the interim analysis for those approaches. The full alpha level is spent at the final analysis.

**REC/FIRST and EX/FIRST**:

Text
$$
f_{\text{OS, rec}}(t) =  \rho_{\text{PFS}}\alpha \cdot 1(t \geq \tau_{\text{OS}}(A_{\text{PFS}})) + \rho_{\text{OS}}\alpha \cdot 1(t \geq 1)
$$

```{r}
#| label: spending_rec_first
os_gs_design_recfirst <- getDesignGroupSequential(
  kMax = 2,
  alpha = alpha,
  sided = 1,
  typeOfDesign = "asUser",
  # Provide user-defined spending function as defined above
  userAlphaSpending = c(
    alpha * bretz_weight_pfs,
    alpha
  ),
  informationRates = c(info_fraction_os, 1)
)
os_stageLevels_recfirst <- os_gs_design_recfirst$stageLevels
os_alphaSpent_recfirst <- os_gs_design_recfirst$alphaSpent
os_alphaIncr_recfirst <- (c(os_alphaSpent_recfirst, alpha) -
  c(0, os_alphaSpent_recfirst))[1:2]
os_criticalValues_recfirst <- os_gs_design_recfirst$criticalValues
```

**REC/GS/LAST and EX/GS/LAST**:

Text
$$
f_{\text{OS, rec}}(t) = 2 - 2 \Phi\left(\frac{\Phi^{-1}(1 - \frac{\rho_{\text{OS}}\alpha}{2})}{\sqrt{t}}\right) + \rho_{\text{PFS}}\alpha \cdot 1(t \geq 1)
$$

```{r}
#| label: spending_rec_gs_last
os_gs_design_recgslast <- getDesignGroupSequential(
  kMax = 2,
  alpha = alpha,
  sided = 1,
  typeOfDesign = "asUser",
  # Provide user-defined spending function as defined above
  userAlphaSpending = c(
    os_alphaSpent_bon[1],
    os_alphaSpent_bon[2] + alpha * bretz_weight_pfs
  ),
  informationRates = c(info_fraction_os, 1)
)
os_stageLevels_recgslast <- os_gs_design_recgslast$stageLevels
os_alphaSpent_recgslast <- os_gs_design_recgslast$alphaSpent
os_alphaIncr_recgslast <- (c(os_alphaSpent_recgslast, alpha) -
  c(0, os_alphaSpent_recgslast))[1:2]
os_criticalValues_recgslast <- os_gs_design_recgslast$criticalValues
```


**REC/GS/FIRST and EX/GS/FIRST**:

Text
$$
f_{\text{OS, rec}}(t) = 2 - 2 \Phi\left(\frac{\Phi^{-1}(1 - \frac{\rho_{\text{OS}}\alpha}{2})}{\sqrt{t}}\right) + \rho_{\text{PFS}}\alpha \cdot 1(t \geq \tau_{\text{OS}}(A_{\text{PFS}}))
$$


```{r}
#| label: spending_rec_gs_first
os_gs_design_recgsfirst <- getDesignGroupSequential(
  kMax = 2,
  alpha = alpha,
  sided = 1,
  typeOfDesign = "asUser",
  # Provide user-defined spending function as defined above
  userAlphaSpending = os_alphaSpent_bon + alpha * bretz_weight_pfs,
  informationRates = c(info_fraction_os, 1)
)
os_stageLevels_recgsfirst <- os_gs_design_recgsfirst$stageLevels
os_alphaSpent_recgsfirst <- os_gs_design_recgsfirst$alphaSpent
os_alphaIncr_recgsfirst <- (c(os_alphaSpent_recgsfirst, alpha) -
  c(0, os_alphaSpent_recgsfirst))[1:2]
os_criticalValues_recgsfirst <- os_gs_design_recgsfirst$criticalValues
```

| Procedure | Local level for OS | Critical value for OS |
| BON | 
| REC/LAST | `{r} alpha` | `{r} round(qnorm(alpha), 4)` |
| REC/FIRST | `{r} alpha` | `{r} round(qnorm(alpha), 4)` |
| EX/LAST | `{r} alpha` | `{r} round(qnorm(alpha), 4)` |
| EX/FIRST | `{r} alpha` | `{r} round(qnorm(alpha), 4)` |
| BON/GS | `{r} round(os_stageLevels_bon[2], 4)` | `{r} round(qnorm(os_stageLevels_bon[2]), 4)` | |
| REC/GS/LAST | `{r} round(os_stageLevels_bon[2], 4)` | `{r} round(qnorm(os_stageLevels_bon[2]), 4)` | |
| REC/GS/FIRST | `{r} round(os_stageLevels_bon[2], 4)` | `{r} round(qnorm(os_stageLevels_bon[2]), 4)` | |
| EX/GS/LAST | `{r} round(inflation_factor_interim * os_stageLevels_bon[2], 4)` | `{r} round(qnorm(inflation_factor_interim * os_stageLevels_bon[2]), 4)` | |
| EX/GS/FIRST | `{r} round(inflation_factor_interim * os_stageLevels_bon[2], 4)` | `{r} round(qnorm(inflation_factor_interim * os_stageLevels_bon[2]), 4)` | |
| OS | `{r} alpha` | `{r} round(qnorm(alpha), 4)` |
### Reallocation from OS to PFS

# Final analysis

## Estimated correlation matrix

## Testing the global null hypothesis

# Summary

# References

::: {#refs}
:::
